# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r}
activityData <- read.csv("activity.csv")
```
## What is mean total number of steps taken per day?
```{r}
## load the plyr library for aggregate functions
library(plyr)

## aggregate the sum of steps by date
steps_by_date <- ddply(activityData,'date',function(activityData) sum=sum(activityData$steps))

## rename the aggregated field to something more descriptive
colnames(steps_by_date)[2] <- "total_steps"

## create a histogram of the total number of steps by date
hist(steps_by_date$total_steps,xlab = "Total Steps by Date", main = "Total Steps by Date")

## calculate the median of the total steps by date
median(steps_by_date$total_steps,na.rm = T)

## calculate the mean of the total steps by date
mean(steps_by_date$total_steps,na.rm = T)
```
## What is the average daily activity pattern?
```{r}
## load ggplot2 library
library(ggplot2)

## group the steps in activity data by intervals for each day
steps_by_interval <- ddply(activityData,'interval',function(activityData) avg = mean(activityData$steps,na.rm = T))

## rename the aggregate column to Average Steps
colnames(steps_by_interval)[2] <- "AverageSteps"

## plot the aggregate data onto a line graph
## with the x axis as intervals and the y axis as average number of steps
qplot(steps_by_interval$interval,steps_by_interval$AverageSteps,geom = "line",main = "Average Steps per Interval", xlab = "Interval", ylab = "Average Number of Steps")

## find the interval that has the highest average number of steps
steps_by_interval[steps_by_interval$AverageSteps == max(steps_by_interval$AverageSteps),]
```
## Imputing missing values
## count the number of NAs for each column in the data
```{r}
## count the number of NAs for each column in the data
count(is.na(activityData))
```
The steps variable has NA values.
```{r}
## create a seperate data set that we can impute
activityData_imp <- activityData

## create a function to intake an interval and output the average steps for that interval
replaceNAs <-
function(x){
    averageSteps <- steps_by_interval[steps_by_interval$interval == x,2]
    return(averageSteps)
}

for(i in 1:length(activityData_imp[,1])){
     if(is.na(activityData_imp[i,1])){
         activityData_imp[i,1] <- replaceNAs(activityData[i,3])
     }
}

## count the number of NAs for each column in the imputted data
count(is.na(activityData_imp))
```
There are no more NA values in the steps variable
```{r}
## aggregate the imputted data to total steps by date
steps_by_date_imp <- ddply(activityData_imp,'date',function(activityData_imp) sum=sum(activityData_imp$steps))

## rename the aggregate column with a descriptive title
colnames(steps_by_date_imp)[2] <- "total_steps"

## create a histogram showing the total steps for each date
hist(steps_by_date_imp$total_steps,xlab = "Total Steps by Date", main = "Total Steps by Date with Imputed Data")
```
Here is the median of the new, imputted data:
```{r}
## find the median of the imputted steps by date data
median(steps_by_date_imp$total_steps,na.rm = T)
```
Here is the mean of the new, imputted data:
```{r}
## find the mean of the imputted steps by date data
mean(steps_by_date_imp$total_steps,na.rm = T)
```
## Are there differences in activity patterns between weekdays and weekends?
```{r}
## convert the date field to the appropriate date data type in R
activityData_imp$date <- as.POSIXct(activityData_imp$date, format="%Y-%m-%d")

## create a weekday factor variable to be added to the dataset
weekday <- factor("weekday",levels = c("weekday","weekend"),labels = c("weekday","weekend"))

## add the weekday variable as a column to the dataset
activityData_imp <- cbind(activityData_imp,weekday)

## assign the weekday variable the appropriate value based on the date field
activityData_imp[weekdays(activityData_imp$date) %in% c("Saturday", "Sunday"),4] <- "weekend"
```
After adding the field and calulating the values for each row, here are the stats:
```{r}
## aggregate the data by weekend vs. weekday and calculate the mean for each level
ddply(activityData_imp,'weekday',function(activityData_imp) avg = mean(activityData_imp$steps))
```
## Here are the plots for weekdays vs. weekends
```{r}
steps_by_interval_week <- ddply(activityData_imp,c('weekday','interval'),function(activityData_imp) avg = mean(activityData_imp$steps,na.rm = T))

## rename the aggregate column to Average Steps
colnames(steps_by_interval_week)[3] <- "AverageSteps"

## create weekday vs. weekend aggregate
plot_weekday <- subset(steps_by_interval_week,weekday == "weekday")
plot_weekend <- subset(steps_by_interval_week,weekday == "weekend")

## create the plot canvas
par(mfrow = c(2,1))
plot(plot_weekend$interval,plot_weekend$AverageSteps,type = 'l',main = 'Weekend',xlab = "Interval", ylab = "Average Steps")
plot(plot_weekday$interval,plot_weekday$AverageSteps,type = 'l',main = 'Weekday',xlab = "Interval", ylab = "Average Steps")